---
title: Rmarkdown introduction
author: "Author: Brendan Gongol"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 4
    fig_caption: yes
    code_folding: hide
    number_sections: true
  pdf_document:
    toc: true
vignette: >
    %\VignetteIndexEntry{Text}
    %\usepackage[utf8]{inputenc}
    %\VignetteEngine{knitr::rmarkdown}
fontsize: 15pt
editor_options: 
  chunk_output_type: console
---

<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: auto !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
body {
text-align: justify}
</style>


# Header 1

## Header 2

### Header 3

#### Header 4

# Header 1


- list 1
- list 2
- list 3

*italics* <br>
**bold** <br>
`emphasize`

Adding a formula centered in the page:
$$A=C+B$$

Adding a formula within the text: $A=C+B$


the following is an example code chunk: 


```{r name1, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
#### Do not install packages in the code elements of the document
# install.packages("ggplot2")
# if (!require("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("ALL")

#### loading packages is OK.
library(ggplot2)
```



```{r, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, class.source = "fold-show", fig.height=5, fig.width=8, results='show', cache=FALSE}
1:4
plot(1:4)
```



executing `rmarkdown::render()` in the Console will knit the markdown file to a pdf or html document.  
















  