---
title: Week 4 Homework Question 1
author: "Author: Laura Wooten"
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`" 
output: 
  html_notebook:
    code_folding: 
    number_sections: true
---
> For this question, you will need to download a data set from the UCI machine learning repository. Separate the data set you downloaded into two groups by the outcome variable that is being studied. In the UCR machine learning repository, most of the time, this will be categorical variable in the “last column” in the data set. If more than one outcome was studied, use the one with the greatest number of records. Then, take a random sample of N=30 measurement from a selected column from each of the subsetted data sets that were partitioned by feature. Then compute the difference in their means and construct a 95% confidence interval for the difference using t.test(). Do this 1,000 times and keep track of all the confidence intervals. One way to answer this question is to use the replicate() function in R, which will return a 2 x 1000 matrix of the lower and upper bound for each of the 1000 replications.

Steps:  
1. read in data
2. acknowledge ? in bare-nuc  
3. Separate data by outcome  
4. Select column of interest  
5. For each df: random sample n=30  
6. difference in sample means -compare against outcomes, 95% CI t-test.  
7. rpt 1000 times

```{r}
data.tbl <- read.table("breast-cancer-wisconsin.data", header = FALSE, sep =",", col.names = c("id", "clump_thickness", "unif_size", "unif_shape", "adhesion", "single_epi_size", "bare_nuclei", "bland_chromatin", "normal_nucleoli", "mitoses", "status"))

#preview
data.tbl[1:5,]
```

```{r}
unlist(lapply(data.tbl[,1:11],class))
#bare_nuclei is character?
unique(data.tbl[,7])

#replace ? with NA and convert to numeric
data.tbl[,7] <- as.numeric(replace(data.tbl[,7], data.tbl[,7]=="?", NA))

#change status column to : 2=Benign, 4=Malignant
data.tbl[,11] <- replace(data.tbl[,11], data.tbl[,11]==2, "Benign")
data.tbl[,11] <- replace(data.tbl[,11], data.tbl[,11]==4, "Malignant")

#change malignant column to factor
data.tbl$status <- as.factor(data.tbl$status)

#separate data into 2 dfs by outcome
ben.data <- data.tbl[data.tbl$status== "Benign",]
mal.data <-  data.tbl[data.tbl$status=="Malignant",]
```

```{r}
#choose a column (independent variable) - clump thickness [,1]
#For each df: random sample n=30  
#difference in sample means -compare against outcomes, 95% CI t-test.
fun1 <- function(ben.data, mal.data, column, smpl.size=30){
  #pull random sample from each outcome
  ben.smpl <- sample(na.omit(ben.data[[column]]), smpl.size, replace = T) #set to replace so that each sample is independent
  mal.smpl <- sample(na.omit(mal.data[[column]]), smpl.size, replace =T)
  
  #calc means
  ben.mu <- mean(ben.smpl)
  mal.mu <- mean(mal.smpl)
  
  #calc difference in means
  diff.mu <- mal.mu - ben.mu #based on quick view of means, malignant outcome is higher
  
  #perform t-test on difference in means
  ci.95 <- t.test(mal.smpl, ben.smpl, conf.level= 0.95)
 
  #return(list(diff.mu, ci.95)) #returns full t-test > try $conf.int
  #return(list(diff.mu, ci.95$conf.int)) #conf.int returns 3 items, try isolate lower and upper bounds only
  return(c(diff.mu, ci.95$conf.int[1], ci.95$conf.int[2]))
}


```
Check first instance (1 simulation) of function, output displayed as difference in sample means (between outcomes), lower bound of 95% confidence interval, higher bound:
```{r}
fun1(ben.data, mal.data, 2)
```
Repeat 1000 times using replicate()
```{r}
ci.matrix <- replicate(1000, fun1(ben.data, mal.data, "clump_thickness"))
#simulations are columns, transpose to make rows
ci.matrix <- t(ci.matrix)
colnames(ci.matrix) <- c("mu_difference", "lower_ci", "upper_ci")
dim(ci.matrix)
#preview
ci.matrix[1:5,]
```

> Then calculate: a. The average length of the confidence interval. b. The population level difference (e.g. the difference between the means) c. How often the confidence intervals contain the population level difference Repeat this analysis for the rest of the columns in the data set and describe the results.

```{r}
fun2 <- function(ci.matrix) {
  # Average length of ci
  ci.length.mu <- mean(ci.matrix[, "upper_ci"] - ci.matrix[, "lower_ci"])
  # Population level difference (average of the differences from each simulation)
  pop.difference <- mean(ci.matrix[, "mu_difference"])
  # How often does the CI include the estimated pop difference?
  popdiff.in.ci <- sum(ci.matrix[, "lower_ci"] <= pop.difference & ci.matrix[, "upper_ci"] >= pop.difference) / nrow(ci.matrix) * 100
  
  return(list(ci_length = ci.length.mu, pop_diff = pop.difference, pct_in_CI = popdiff.in.ci))
}

fun2(ci.matrix)
```
Repeat for all columns:

```{r}
#data frame to store stats for each column
column.stats <- data.frame(column_name = character(), avg_ci_length= numeric(), pop_level_difference= numeric(), percent_in_ci= numeric())

#create empty list to store sample data for later
sample.matrices <- list()

for (column.i in colnames(data.tbl)[2:10]){
  col.ci.matrix <- replicate(1000, fun1(ben.data, mal.data, column.i))
  col.ci.matrix <- t(col.ci.matrix)
  colnames(col.ci.matrix) <- c("mu_difference", "lower_ci", "upper_ci")
  
  #store matrix in the list for later use (to compare against population data)
  sample.matrices[[column.i]] <- col.ci.matrix
  
  #calculate summary stats, add to df
  calculations <- fun2(col.ci.matrix)
  row.df <- data.frame(column_name= column.i,avg_ci_length= calculations$ci_length, pop_level_difference= calculations$pop_diff, percent_in_ci= calculations$pct_in_CI)
  #bind new row to column.stats
  column.stats <- rbind(column.stats, row.df)
}
column.stats
```
**DESCRIBE RESULTS**  
These results depict the differences in measurements between benign and malignant tissue samples from the breast cancer data set. The confidence intervals represent a measure of reliability of our statistical analysis. The average length of the confidence intervals varied across each feature, with bland_chromatin producing the most precise estimate (smallest spread) and bare_nuclei showing the widest spread, suggesting greater variability in its measurements. The population-level difference, which is the estimated mean difference between benign and malignant cases, was largest for bare_nuclei at 6.25, which potentially indicates a strong association with malignancy, whereas mitoses at 1.55 exhibited a much smaller difference, therefore meaning a weaker predictive relationship between variables. Across the variables, the confidence intervals contained the population-level difference approximately 95% of the time, aligning with expectations as we constructed a linear model from a confidence level of 0.95. This suggests that the statistical analysis is well constructed, and can provide a reliable estimate of the mean differences. However, mitoses (91.7%) showed a slightly lower coverage percentage, which may reflect greater variability in the data. Looking at these results as a whole, features with large mean differences and narrow confidence intervals show stronger separation between groups, whereas those with smaller differences and wider intervals may indicate weaker predictive reliability.  

---

>Next, calculate the confidence interval of the entire set of samples and compare it to by the randomly samples confidence intervals. A visual representation might be nice here which illustrates the confidence interval of the entire dataset relative to the randomly sampled confidence intervals

```{r eval=FALSE, include=FALSE}
## NA - CHANGED HOW I STRUCTURED THIS----- SEE CODE CHUNCK BENEATH
#__________________________________________________
#t.test each entire column, compare to samples. 
for (i in 1:nrow(column.stats)){
  
  #pull sample data from column.stats
  column.i <- column.stats$column_name[i]
  pop.diff <- column.stats$pop_level_difference[i]
  avg.length <- column.stats$avg_ci_length[i]
  
  #perform t-test on entire columns
  full.ttest <- t.test(mal.data[[column.i]], ben.data[[column.i]], conf.level = 0.95)
  full.ci <- full.ttest$conf.int
  full.diff <- full.ttest$estimate[1] - full.ttest$estimate[2] #estimate pulls means from varX(malignant) vary(benign)
  
  #plot

}
```

```{r}
pop_vs_sample_plots <- function(sample.matrices, ben.data, mal.data, column){
  #assign correct 1000x3 matrix for column from the list
  tmp.matrix <- sample.matrices[[column]]

  
  #t-test entire column (population data)
  full.ttest <- t.test(mal.data[[column]], ben.data[[column]], conf.level = 0.95)
  full.ci <- full.ttest$conf.int
  full.diff <- full.ttest$estimate[1] - full.ttest$estimate[2]  #estimate pulls means from varX(malignant) vary(benign)
  
  #define plot limits
  y_range <- range(c(tmp.matrix[, "lower_ci"], tmp.matrix[, "upper_ci"], full.ci)) #could also use max?
  plot(1:1000, 1:1000, type = "n",
       xlab = "n 1000 sims", ylab = "Mean Difference",
       ylim = y_range,
       main = paste("Simulated Confidence Intervals for", column))
  #plot each iid's confidence interval as a vertical line
  for (i in 1:1000) {
    segments(x0 = i, x1 = i, 
             y0 = tmp.matrix[i, "lower_ci"], 
             y1 = tmp.matrix[i, "upper_ci"],
             col = rgb(0, 0, 1, alpha = 0.3))
  }
  
  # add each i.i.d's mean differences as points.
  points(1:1000, tmp.matrix[, "mu_difference"], pch = 16, col = rgb(0, 0, 1, alpha = 0.5))
  
  #lines for the bounds of the population CI
  abline(h = full.ci[1], col = "red", lty = 2, lwd = 2)
  abline(h = full.ci[2], col = "red", lty = 2, lwd = 2)
  #line for the population mean diff
  abline(h = full.diff, col = "green", lwd = 2)
  
  # add a legend
  legend("topright",
         legend = c("Difference in smpl means", "Sample CI's", "Population CI", "Difference in pop mean"), col = c(rgb(0, 0, 1, alpha = 0.5),rgb(0, 0, 1, alpha = 0.3), "red", "green"),lty = c(NA, 1, 2, 1), pch = c(16, NA, NA, NA), cex=0.5)
}
  

```

```{r}
#perform function on each column
for (column.i in colnames(data.tbl)[2:10]){
  pop_vs_sample_plots(sample.matrices,ben.data,mal.data, column.i)
}
```
I tried to put all these plots together but they ended up being too small. Each blue vertical line represents a confidence interval from a single random sample of N=30, while the blue dots indicate the mean difference for each sample. The red dashed lines mark the upper and lower bounds of the confidence interval from the entire population, while the green line represents the population mean difference. Each of these plots shows the variability in the confidence intervals from repeated samples. The spread of the individual sample confidence intervals shows how much estimates fluctuate from sampling variation. The population confidence interval provides a standard to compare our samples to, showing where the true underlying difference lies. However, I understand that this ‘population’ is still a sample from the broader underlying population, and while we treated it as a reference, it does not represent the absolute population but is an estimate itself based on the available data from the set. If a large proportion of the individual confidence intervals contain the population mean difference, it supports our sampling approach produces reliable estimates even under the limitations of working with a finite dataset.
